\RequirePackage[log]{snapshot}
\documentclass[twocolumn]{svjour3}

% Basic packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{mathptmx}   % use Times fonts if available
\usepackage{amsmath,amssymb}
\usepackage{graphicx}

% More formating
\usepackage[ruled,vlined]{algorithm2e}
%\usepackage{comment}
\usepackage{mynotations}
%\usepackage{mystyle}

% Other packages
\usepackage{url}
\usepackage[colorlinks=true]{hyperref}   % last package

% Some settings
\journalname{Journal of Mathematical Imaging and Vision}
\smartqed
\graphicspath{{./images/}}

% Some commands
\newcommand{\rednote}[2][Note]{\medskip\noindent\textbf{\color{red}#1} --- #2\medskip}
\newcommand{\todo}[1]{\rednote[To Do]{#1}}

% Start
\begin{document}


%%%%%%%%%%%%%%%%%%%%  Header of the paper  %%%%%%%%%%%%%%%%%%%%

\date{Received: date / Accepted: date}   % entered by the editor

\title{Variational Texture Synthesis with Sparsity and Spectrum Constraints}
%\subtitle{Complementarity of Patch Sparsity and Spectrum Constraints}
%\titlerunning{Variational Texture Synthesis}

\authorrunning{G.~Tartavel, Y.~Gousseau, and G.~Peyr\'e}
\author{
  Guillaume Tartavel \and
  Yann Gousseau \and
  Gabriel Peyr\'e
}

\institute{
  Guillaume TARTAVEL \at
Telecom ParisTech / LTCI CNRS \\
    46 rue Barrault -- F-75634 Paris Cedex 13 -- France \\
    Tel.: +33 1 45 81 73 91 \\
    \email{tartavel@telecom-paristech.fr}
  \and
  Yann GOUSSEAU \at
Telecom ParisTech / LTCI CNRS \\
    46 rue Barrault -- F-75634 Paris Cedex 13 -- France \\
    Tel.: +33 1 45 81 78 41 \\
    \email{gousseau@telecom-paristech.fr}
  \and
  Gabriel PEYR\'E \at
    CNRS / Ceremade, Universit\'e Paris-Dauphine \\
    Pl.~Mar\'echal De Lattre De Tassigny -- 75775 Paris Cedex 16 -- France \\
    Tel.: +33 1 44 05 48 71 \\
    \email{peyre@ceremade.dauphine.fr}
}

\maketitle


%%%%%%%%%%%%%%%%%%%%  Content of the paper  %%%%%%%%%%%%%%%%%%%%

\begin{abstract}

This paper introduces a new approach for texture synthesis.
We propose a unified framework that both imposes first order statistical constraints on the use of atoms from an adaptive dictionary, as well as second order constraints on pixel values.
This is achieved thanks to a variational approach, the minimization of which yields local extrema, each one being a possible texture synthesis.
On the one hand, the adaptive dictionary is created using a sparse image representation rationale, and a global constraint is imposed on the maximal number of use of each atom from this dictionary.
On the other hand, a constraint on second order pixel statistics is achieved through the power spectrum of images.
In a sense, this method reconciles the first and second theory proposed by B.~Julesz for texture discrimination.
An advantage of the proposed method is its ability to truly synthesize textures, without verbatim copy of small pieces from the exemplar.
In an extensive experimental section, we show that the resulting synthesis achieves state of the art results, both for structured and small scale textures.

\keywords{
  texture synthesis \and
  variational methods \and
  statistical constraints \and
  Fourier spectrum \and
  sparse decomposition \and
  dictionary learning \and
  random phase textures \and
  Gaussian textures
}

% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}


% Sections

\input{sections/sec-introduction}
\input{sections/sec-variational}
\input{sections/sec-framework}
\input{sections/sec-results}
\input{sections/sec-conclusion}
\input{sections/app-proofs}
%\input{sections/app-projections}   % for our comprehension only


%%%%%%%%%%%%%%%%%%%%  End of the paper  %%%%%%%%%%%%%%%%%%%%

\begin{acknowledgements}
  We would like to thanks the authors of the VisTex database~\cite{pickard1995vistex} for the set of textures they publicly provide.
  Gabriel Peyré also acknowledges support from the European Research Council (ERC project SIGMA-Vision).
  Yann Gousseau acknowledges support from the French National Research Agency (ANR pro\-ject MATAIM).
\end{acknowledgements}

\bibliographystyle{spmpsci}
\bibliography{bibliography}

\end{document}
